---
title: "MATH70076: Data Science - Formative Assessment"
subtitle: 'MSc in Statistics 2026/27, Imperial College London'
author: "02398379"
format:
  html:
    toc: true
    highlight: tango
    self-contained: true
  pdf: default
format-links: false
bibliography: example.bib 
---

**Deadline:  Monday 12 October 2026 at 13:00.**

_For this formative assessment you should submit two files via the Imperial College VLE on Canvas by the deadline stated above. Your files should be named as follows:_

- `YOURCID-MATH70076-formative-assessment.pdf`: your rendered report,
- `YOURCID-MATH70076-formative-assessment.zip`: a zip file containing the relevant source code to generate your report.

_All submitted materials should be clearly presented and be understandable as stand-alone documents._

_Please note that large files can take quite some time to upload. Ensure that you upload each document to the correct part of the learning space in a timely manner._

_This mock coursework is a formative assessment and so will not contribute to your overall grade for the module. It is designed for you to practice the skills you have learned in the first workshop and to get feedback on your reflective writing style. This activity is expected to take approximately 6 hours of individual effort. Assessment criteria are given in the "set yourself up for success" boxes. Satisfying 18 or more out of these 23 criteria corresponds to a notional "pass" grade._

_In submitting this mock assessment you certify that it is entirely your own work, apart from where otherwise acknowledged, and includes no plagiarism. Note that software tools are used as part of plagiarism detection._

-----

## Background 

### Generalised Pareto Distribution 

The Generalized Pareto Distribution (GPD) is a flexible family of continuous probability distributions that arises naturally in extreme value theory, particularly for modelling the distribution of excesses over a threshold. It is parametrised by a shape parameter $\xi \in \mathbb{R}$, a scale parameter $\sigma > 0$, and a location parameter $u \in \mathbb{R}$. Its cumulative distribution function (CDF) is given by

$$
F(x;\,\sigma,\xi, u) = 
\begin{cases}
1 - \left(1 + \dfrac{\xi (x-u)}{\sigma}\right)_+^{-1/\xi}, & \xi \neq 0, \, x \geq u; \; \\[1.2em]
1 - \exp\!\left(-\dfrac{x-u}{\sigma}\right), & \xi = 0, \, x \geq u;
\end{cases}
$$ {#eq-gpd-cdf}

where $x_+ = \max(x,0)$ and its probability density function (PDF) is

$$
f(x;\sigma,\xi,u) =
\begin{cases}
\dfrac{1}{\sigma}\left(1 + \dfrac{\xi (x-u)}{\sigma}\right)_+^{-1/\xi - 1}, & \xi \neq 0, \\[1.2em]
\dfrac{1}{\sigma}\exp\!\left(-\dfrac{x-u}{\sigma}\right), & \xi = 0,
\end{cases}
$$ {#eq-gpd-pdf}


defined on the same support as the CDF. The GPD encompasses a variety of tail behaviours:

- when $\xi > 0$ the GPD has heavy, slowly decaying tails, 
- in limiting case when $\xi \rightarrow 0$ the GPD reduces to an exponential distribution;
- when $\xi < 0$ the GPD has light, quickly decaying tails with a finite upper endpoint of $x^+ = u - \sigma / \xi$. 

### Probability Integral Transform

The probability integral transform states that if a random variable $X$ has a continuous cumulative distribution function $F_X(x)$, then the transformed variable $A = F_X(X)$ follows a uniform distribution on $[0,1]$. Conversely, if $F$ is an invertible function then $Y = F^{-1}_X(A)$ has the same distribution as $X$.

## Questions 

###  Question 1 

Derive an expression for the inverse cumulative distribution function (also known as the quantile function) $F^{-1}_X: [0,1] \rightarrow [u, x^+]$ of $X \sim \text{GPD}(u, \sigma, \xi)$. Your answer should refer to at least one equation given in the background material. 

::: {.callout-tip}
## Set yourself up for success

Does your answer contain:

- a few sentences of text describing your approach to the problem; 
- a reference to at least one equation from the background section;
- a correctly formatted LaTeX equation;
- a valid approach to the problem and correct expression.
:::


Since $F_X$ from @eq-gpd-cdf is continuous, then $A = F_X(X)$ with $A \sim U[0,1]$ by the probability integral transform. Therefore, let $p \in (0,1)$, the inverse CDF can be obtained by setting $p = F_X(x)$, and solve for $x$. Since @eq-gpd-cdf has separate forms for $\xi \neq 0$ and $\xi = 0$, we will considered the two cases separately. For each cases, we will also consider the endpoints $p=0$ and $p=1$ seperately.

If $\xi \neq 0$, then

$$
\begin{align}
  p &= 1 - \left(1 + \dfrac{\xi (x-u)}{\sigma}\right)_+^{-1/\xi} \iff\\
  \left(1 + \dfrac{\xi (x-u)}{\sigma}\right)_+^{-1/\xi} &= 1-p  \quad \iff \\
  \left(1 + \dfrac{\xi (x-u)}{\sigma}\right)^{-1/\xi} &= 1-p  \quad \text{since } 1-p >0 \quad \iff \\
  1 + \dfrac{\xi (x-u)}{\sigma} &= (1 - p)^{-\xi}  \quad \iff \\
  x &= u + \dfrac{\sigma[(1 - p)^{-\xi}-1]}{\xi} \; .
\end{align}
$$

If $\xi = 0$, then

$$
\begin{align}
  p &= 1 - \exp\!\left(-\dfrac{x-u}{\sigma}\right) \iff\\
  \exp\!\left(-\dfrac{x-u}{\sigma}\right) &= 1-p \quad \iff \\
  -\dfrac{x-u}{\sigma} &= \log(1-p) \quad \iff \\
  x &= u - \sigma\log(1-p) \; .
\end{align}
$$

Now if $p = 0$, then the expressions are valid in both cases since $1-p > 0$, and we have $F^{-1}_X(0) = u$. However, consider $p=1$, then we have issues. If $\xi < 0$, we get $F^{-1}_X(0) = u-\sigma/\xi = x^+$; if $\xi > 0$, then $(1 - p)^{-\xi} \rightarrow +\infty$, and if $\xi = 0$, then $\log(1-p) \rightarrow -\infty$, so for $\xi \geq 0$, we get $F^{-1}_X(1) = +\infty$.

Altogether, the inverse CDF of $X \sim \text{GPD}(u,\sigma,\xi)$, is given by

$$
F_X^{-1}(p;\,\sigma,\xi, u) = 
\begin{cases}
u + \dfrac{\sigma[(1 - p)^{-\xi}-1]}{\xi}, & \xi \neq 0, \, p \in [0,1); \; \\[1.2em]
u - \sigma\log(1-p), & \xi = 0, \, p \in [0,1); \; \\[1.2em]
u-\sigma/\xi , & \xi < 0, \, p = 1; \; \\[1.2em]
+\infty , & \xi \geq 0, \, p = 1.
\end{cases}
$$ {#eq-gpd-qf}


### Question 2 

_For this question you should display the R code you use to define and document `qgpd()` within the main text of your report._

(a) Write and document your own function `qgpd()` to calculate quantiles of a given generalised Pareto distribution. Your function should have inputs and behaviour similar to the built-in R functions such as `qnorm()` and `qunif()` and check that inputs are in the correct format. 

(b) Suppose a random variable $X$ follows an generalised Pareto distribution with threshold parameter $u=1.5$, scale parameter $\sigma = 2$ and shape parameter $\xi = -0.4$. Use your function to find quantiles $x_p$ for $p = 0.5, 0.75, 0.99$ (i.e. for each $p$ find the value for which $\Pr(X < x_p) = p$).


:::{.callout-tip}
## Set yourself up for success

Does your answer: 

- contain a valid R function definition;
- document the expected inputs, outputs and behaviours of that function;
- check the validity of inputs;
- handle any edge-cases in an appropriate way;
- return the correct quantile values? 

:::


(a)

```{r}
qgpd <- function(p, u, sigma, xi){
  # Check that probabilities are between 0 and 1
  if (any(p < 0 | p > 1, na.rm = TRUE)) {
    stop("p must be between 0 and 1.")
  }
  
  # Check that u is a single value, not a vector
  if (length(u) != 1) {
    stop("u must be a single real value.")
  }
  
  # Check that sigma is a single positive value
  if (length(sigma) != 1 || sigma <= 0) {
    stop("sigma must be a single positive real value.")
  }
  
  # Check that xi is a single value
  if (length(xi) != 1) {
    stop("xi must be a single real value.")
  }
  
  # Calculate the quantiles using the inverse formula from Q1
  if (xi != 0) {
    u + sigma*((1 - p)^(-xi) - 1)/xi
  } else {
    u - sigma * log(1 - p)
  }
}
```

The function `qgpd()` takes four inputs. The argument $\texttt{p}$ is a vector of probabilities between $0$ and $1$; $\texttt{u}$, $\texttt{sigma}$ and $\texttt{xi}$ are the location, scale and shape parameters of the GPD, respectively. The parameters $\texttt{u}$ and $\texttt{xi}$ are each required to be a single real value, and $\texttt{sigma}$ must be a single positive real value. 

If any probability lies outside the interval $[0,1]$, or if any of the parameters do not satisfy the required conditions, the function returns an error message. Otherwise, the function returns a vector of quantiles corresponding to the probabilities of $\texttt{p}$. Missing values in $\texttt{p}$ are preserved as missing values in the output in the same way as to `qnorm()` and `qunif()`.

(b)

Let $X \sim \text{GPD}(1.5, 2, -0.4)$, and $\Pr(X < x_p) = p$. Using part (a), we get
```{r}
#| echo: FALSE

p = c(0.5, 0.75, 0.99)
x_p = qgpd(p, 1.5, 2, -0.4)

quantiles <- data.frame(p, x_p)
knitr::kable(x = round(quantiles, 2), col.names = c("$p$", "$x_p$"))
```


### Question 3

_For this question, all R code should be displayed only within the appendix, not in the main report._

The file `gpd_samples.csv` contains six sets of random variates generated from different generalised Pareto distributions. The details of the generalised Pareto distributions used are summarised in `gpd_parameters.csv`. Unfortunately some of the parameter sets were recorded incorrectly. 

Within a single figure, construct a series of quantile-quantile plots to identify which datasets are inconsistent with their stated distributions. You should both justify your conclusions and describe your level of confidence in your findings. 

::: {.callout-tip}
## Set yourself up for success 

Does your solution contain: 

- at least one quantile-quantile plot;
- six qq-plots in a single figure;
- use of loops, vectorisation, or function definitions to avoid repetitive code.
- figures with clear text, useful captions and appropriate visual mapping of data;
- a few paragraphs describing and justifying your findings and referencing the figure;

:::

```{r}
#| echo: FALSE

gpd_samples <- read.csv("gpd_samples.csv")
gpd_parameters <- read.csv("gpd_parameters.csv")
```

We obtain the samples and the parameters from the files `gpd_samples.csv` and `gpd_parameters.csv`, and use them to obtain qq-plots for each of the 6 pairs of generalised Pareto distribution.

```{r}
#| echo: FALSE
#| layout-ncol: 3
#| label: fig-qqplots
#| fig-cap: "QQ-plots comparing the observed and theoretical quantiles of the recorded GPD."
#| fig-subcap: 
#|   - "QQ-plot for GDP a"
#|   - "QQ-plot for GDP b"
#|   - "QQ-plot for GDP c"
#|   - "QQ-plot for GDP d"
#|   - "QQ-plot for GDP e"
#|   - "QQ-plot for GDP f"

for (i in 1:6) {
  id <- gpd_parameters$id[i]
  sample_val <- sort(gpd_samples$value[gpd_samples$set_id == id])

  p <- ppoints(length(sample_val))
  u <- gpd_parameters$u[i]
  sigma <- gpd_parameters$sigma[i]
  xi <- gpd_parameters$xi[i]

  theoretical_val <- qgpd(p, u, sigma, xi)

  plot(theoretical_val, 
       sample_val, 
       xlab = "Theoretical quantiles", 
       ylab = "Sample quantiles", 
       main = paste("QQ-plot for GDP", id))
  
  abline(a = 0, b = 1, lty = 2)
}

```

@fig-qqplots suggests that samples $\textbf{a}$ and $\textbf{d}$ are inconsistent with their recorded parameters. For sample $\textbf{a}$, the points in the upper tail of @fig-qqplots-1 drifts away from the dashed reference line $y=x$. Also, since its recorded shape parameter is $\xi=-0.4 < 0$, the distribution must has finite upper endpoint $x^+=u-\sigma/\xi=6.5$, but the sample clearly contains an observation greater than $8$. Therefore, there are some strong evidences that suggests the parameters recorded for sample $\textbf{a}$ are incorrect. That aside, in @fig-qqplots-4, qq-plot for sample $\textbf{d}$ are in general, completely below the reference line, which indicates that its recorded parameters are incorrect.

In contrast, sample $\textbf{b}$ follows the reference line closely in @fig-qqplots-2. Samples $\textbf{c}$, $\textbf{e}$ and $\textbf{f}$ has some deviations in their upper tails, shown on @fig-qqplots-3, @fig-qqplots-5, and @fig-qqplots-6, respectively, but these are weak evidence of incorrect parameters. Sample $\textbf{c}$ only has 20 observations, which is insufficient to deduce a significant conclusion. Samples $\textbf{e}$ and $\textbf{f}$ both have recorded $\xi >0$ and hence a heavy upper tail fits with the plots. Overall, I would suggest to only identify $\textbf{a}$ and $\textbf{d}$ as the incorrectly recorded parameter sets.

### Question 4

_For this question, all R code should be displayed only within the appendix, not in the main report._

A hydrologist is interested in understanding the river flow at a location that is historically prone to flooding. There is a river flow gauge nearby which measures river flow in units of cubic meters per second ($\text{m}^3/\text{s}$ or cumecs). Based on her knowledge of other rivers, the hydrologist proposes that for this river gauge:

 - the distribution of river flow values is constant over time 
 - for river flows exceeding 75 cumecs, it is appropriate to model these data as independent and identically distributed GPD($\sigma =29.7$, $\xi=0.62$, $u = 0$). 

Use `riverflow_2015-2024.csv` to conduct an exploratory investigation of whether these proposals are valid for the dataset provided. Summarise your findings in 250-350 words, supporting these with a collection of 4 visualisations/figures. 


::: {.callout-tip}
## Set yourself up for success 

Does your answer contain: 

- 400-500 words of text clearly describing your choice of visualisations, their interpretation and your conclusions about the validity of each assumption,
- A series of 4 visualisations / figures,
- A varied and appropriate choice of figures to investigate each of the hydrologist's proposals,
- Figures with clear text, useful captions and appropriate visual mapping of data;
- At least one reference to each visualisation within the main text.
:::


<!-- YOUR INVESTIGATION GOES HERE --> 



::: {.callout-tip}
## Set yourself up for success 

- Does your document render without any formatting issues? 
:::

## Question 5 Reflective Summary 

Describe, evidence and reflect on the process of constructing the visualisations in your answer to Question 4. 

Your response should be 300-400 words and might, for example, comment on:

- why you chose these forms of visualisation and what alternatives you considered; 
- a challenge you faced or something that you learned while making the visualisations; 
- how you might approach this task differently with more time or for a different intended audience. 

::: {.callout-tip}
## Set yourself up for success 

Does your response: 

- describe how you engaged with the visualisation activity;
- give specific evidence of that engagement;
- critically reflection on that engagement?
:::

_End of Assessment._

# Code Appendix {#sec-code-appendix}

```{r ref.label=knitr::all_labels()}
#| echo: true
#| eval: false
#| code-fold: true
```






